Highly Available Transactions: Virtues and Limitations

### 2013-03-26 writeup/hat-analysis.tex
Another pass at writing

When possible, we draw on existing properties and definitions from the
database and distributed systems literature, providing a brief,
informal explanation of each guarantee and illustrate it with an
example. In particular, for our ACID guarantees, we draw largely on
Adya's dissertation~\cite{adya}, Berenson et al.'s 1995 critique of
the ANSI SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in the extended version of this work.

@@

To begin, we present achievable semantics, offering proof-of-concept
algorithms to demonstrate feasibility.

@@

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and Berenson's ``broad'' $P1$
(2.2)). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. Clients can buffer their writes until
they commit, or, alternatively, can send them to servers, who will not
serve new writes until clients notify them that the writes have been
committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are promised by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatabable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read \texttt{null} for each data item. However, when coupled with
additional properties, like transactional atomicity, it is stronger:
the repeated reads must obey additional ordering constraints. In fact,
distributed systems describe a ``consistent snapshot'' across a set of
related events as a \textit{cut} across a set of participants or, in
our case, data items. Accordingly, to capture the notion that the
transaction should read from a (non-changing) consistent cut over data
items (and to disambiguate from the aforementioned stronger Repeatable
Read properties), we call ANSI SQL Repeatable Read (preventing P2)
\textbf{Cut Isolation}.It is possible to satisfy cut isolation with
high availability by caching appropriate versions for a transaction's
reads or, alternatively, using multi-versioning on the server and
ensuring that each transaction's successive reads return the same
version of each data item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items plus logical ranges of data
items, or predicate-based reads. This prevents the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}. We call predicate-based cut isolation, which prevents
Phantoms, \textit{Phantom Cut Isolation}.\footnote{Oracle provides an
  isolation level called ``Statement Level Read Consistency''; this is
  analogous to Phantom Cut Isolation at the level of a single
  statement. We do not consider this isolation model here as it is
  subsumed by standard Snapshot Isolation and we believe our
  discussion is easily extended to incorporate it.}

@@

Transactional atomicity is core to ACID guarantees. Although, at least
by the acronym, it is not an ``isolation'' property, transactional
atomicity restricts transactions' ability to view the effects of
partially completed transactions. Under transactionational atomicity,
within each transaction, either all effects of another transaction are
observed, or none are (equivalently, once some of the effects of a
transaction are observed, all effects are observed). This is a
strictly stronger guarantee than Read Committed isolation: in Read
Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. For example, if
$T_1$ commits, then, given transactional atomicity, $T_2$ must observe
$b=c=1$ (or later versions):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$.

Perhaps perplexingly, discussions of transactional atomicity are
absent from existing discussions of weak isolation levels. This is
perhaps again due to the single-node context in which prior work was
developed: on a single server, atomicity is near-trivial to achieve
via lightweight locking and/or local concurrency control over data
items. (In this spirit, multi-item atomicity is common in so-called
``entity groups,'' which logically represent groups of co-located
items.) In contrast, in a distributed environment, atomicity over
arbitrary groups of items is considerably more difficult to achieve
with high availability. The key challenge is that different servers
responsible for subsets of a transaction's updates must coordinate
regarding when to reveal their writes to a transaction---without
violating availability due to, say, locking, or a missing
transactional write. Together with Cut Isolation, Transactional
Atomicity prevents Read Skew anomalies (Berenson et al.'s
A5A~\cite{ansicritique}).

Transactional atomicity is achievable in a HAT environment. To ensure
transactional atomicity, servers only need to know when all of a
transaction's updates are present on their respective servers but do
not need to know when all respective servers know when all of a
transaction's updates are present on their respective servers. This
subtle distinction corresponds to the difference between the
distributed systems concepts of consensus and reliable broadcast. The
former requires explicit agreement across servers, while the latter
simply requires that, if one server reveals a write, then all servers
reveal the write. We omit a full discussion of the algorithm and
instead refer the reader to the extended version of this paper.

@@

So far, our models have not considered guarantees across transactions,
but, in the distributed systems context, many useful \textit{safety}
guarantees span multiple operations. There is no guarantee of forward
progression between transactions or any guarantees on the ordering
between transactions (other than that there exists some arbitrary
ordering). \textit{Session guarantees} are used to describe continuity
between groups of transactions issued by the same logical
user. Informally, a \textit{session} is a form of context that
consists of logical state that should persist between
operations. Users accessing a web site may want to experience a notion
of ``forward progress'' for the duration that they are logged in, so
all transactions submitted between ``log in'' and ``log out'' might
form a session. We can define sessions at several granularities (e.g.,
per-transaction, per-client, per-hour) but, for simplicity, will refer
to client-level guarantees in the discussion.

@@

\vspace{.25em}\noindent\textit{\textbf{Monotonic reads}} stipulates
that a client's subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.25em}\noindent\textit{\textbf{Monotonic writes}} requires
that transactions from each individual client be serialized. Any order
on transactions should also respect the order in which each session
submitted transactions.

\vspace{.25em}\noindent\textit{\textbf{Writes Follow Reads}} requires
that, if a session reads a write originating from transaction $T_1$
then performs transaction $T_2$, then a session can only read $T_2$'s
writes if it can also read $T_1$'s prior write (or overwritten values
for $T_1$'s write). This requirement forms the basis for potential
causality: writes follow reads obeys the ``reads-from'' relation that
captures all events that influenced each
transaction~\cite{causalmemory}. Any transaction order should respect
the reads-from order (note that the total order on transactions
precludes reads-from cycles).

@@

\noindent\textit{\textbf{Read your writes}} requires
that whenever a client reads a given data item after updating it, the
read returns the updated value (or a value with a higher ID).

\noindent\textit{\textbf{PRAM}} provides the illusion of serializing
each session's operations and is the combination of monotonic reads,
monotonic writes, and read your writes.

\noindent\textit{\textbf{Causal consistency}} is the combination of
all of the session guarantees~\cite{sessiontocausal} (alternatively,
PRAM and writes-follow-reads) and is also referred to as PL-2L
isolation~\cite{adya}).

@@

A HAT system can make \textbf{limited consistency} guarantees. It can
execute commutative and logically monotonic~\cite{needed} operations
without the risk of invalidating (also monotonic) application-level
integrity constraints. Our goal in this paper is not to sketch the
entire space of consistency models that are achievable (see
Section~\ref{sec:futurework}, however we specifically evaluate TPC-C
transaction semantics under HAT consistency guarantees in
Section~\ref{sec:evaluation}.

As we briefly discussed in Section~\ref{sec:availability}, a client
requiring \textbf{durability} of surviving $F$ server faults requires
at least ($F+1$)-availability.

Finally, to require that servers propagate writes between replicas, we
can require \textbf{convergence}, or eventual consistency for each
data item: in the absence of new mutations to a data item, in the
absence of partitions, all servers eventually agree on the value for
each item. This is typically accomplished by any number of
anti-entropy protocols, which periodically update neighboring servers
with the latest value for each data item. This convergence property
should be considered mandatory given the spirit of existing literature
on isolation protocols: in all models we have encountered, the
database should behave as if there is some total order over
transactions. Indeed, non-convergent systems are remarkably difficult
to reason about.

@@

While there are infinitely many HAT models
(Section~\ref{sec:futurework}), at this point, we have largely
exhausted the range of achievable, previously defined (useful) HAT
semantics. Before summarizing our possibility results, we will present
impossibility results for HATs, also defined in terms of previously
identified isolation and consistency anomalies.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and Serializability---requires unavailability.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)
\end{align*}
The final state of the database will converge to an execution that
could not have resulted from a serial excecution. To prevent this from
happening, either $T_1$ or $T_2$ should not have committed. Each
client's respective server might try to detect that another write
occurred, but this requires knowing the version of the latest write to
$x$. In this example, this reduces to a requirement for
linearizability, which is, via Gilbert and Lynch's proof of the CAP
Theorem, provably unachievable with high availabilty.

@@

Data storage engines make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability, which states that reads will return the last
completed write to a data item, and there are several other (weaker)
variants such as safe and regular register semantics. When applied to
transactional semantics, the combination of one-copy serializability
and linearizability is called \textit{strong (or strict) one-copy
  serializability} or \textit{externally consistent transactions}
(e.g., Google's Spanner~\cite{spanner}). It is also common,
particularly in systems that allow reading from masters and slaves, to
provide a guarantee such as ``read a version that is no more than five
seconds out of date'' or similar. Unfortunately, an indefinitely long
partition can cause violation of any recency bound.

@@

We illustrate the hierarchy of available, sticky available, and
unavailble consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable: for example, a HAT system can provide monotonic reads and
transactional atomicity or transactional atomicity and read your
writes and, for maximum semantic richness, can provide all of
Transactional Atomicity, Predicate Cut Isolation, Monotonic Reads,
Monotonic Writes, and Writes Follow Reads. An unavailable system might
provide PRAM and Serializability, as advocated by Salem and
Daudjee~\cite{needed}.

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Perhaps satisfactorily, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., only considering client-centric session
consistency guarantees, the diagram depicts 32 possible HAT
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonimization is \textit{not} exhaustive, but we
believe it lends substantial clarity into the relationships between a
large subset of the prominent ACID and distributed consistency
moels. Additional models of read/write transactions that we have
omitted should be easily classifiable based on the available
primitives and unavailable anomalies we have already discussed.

@@

\begin{table}
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, Conflict Serializability
(CSR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
\\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.8]
  \tikzstyle{sticky}=[rectangle,draw=blue!50,fill=blue!20,thick]
  \tikzstyle{noha}=[ellipse,draw=red!50,fill=red!20,thick, inner sep=0pt,minimum size=12pt]

@@

\noindent\textbf{Model Composition} may be non-trivial. Consider the following transactions:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)
\end{align*}
If we want to guarantee both cut isolation and transactional atomicity
and these are the only transactions that are executed by the system,
then $T_3$ needs to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means
that either the implementation should frequently return null
(definitely undesirable and possibly non-convergent), keep multiple
versions of each data item (necessitating potentially complicated
distributed garbage collection), or use pre-declared read sets to
fetch a consistent cut of keys before the transaction begins to
execute. Using client-side caching can alleviate some of these
challenges, but then the system becomes sticky high available.

Composition cost also varies by property. For instance, Charron-Bost
has proven that, to capture causality between $N$ communicating
processes, standard vector-based approaches face an upper bound of
$O(N)$~\cite{charron-bost} storage per write. This means that, with
$100,000$ clients, each write might be accompanied by $100,000$
timestamps per vector. This is difficult to scale. By compromising on
availability (e.g., treating a datacenter as a linearizable cluster),
this overhead can be reduced~\cite{cops, eiger}, but it is much
cheaper to provide, say, read your writes, compared to causality.

\noindent The relationship between \textbf{Transactional Atomicity and
  Linearizability} is non-obvious. One might attempt to find a mapping
between the two: Transactional Atomicity dictates that writes to
multiple keys across multiple servers are made visible to readers all
at once, while Linearizability dictates that writes to a single key on
multiple servers are made visible to all readers at once. The
difference is two-old. First, in linearizable (and safe and regular)
systems, writes are made visible to all clients \textit{immediately}
after they finish. With Transactional Atomicity, there is no recency
requirement. Second, in linearizabile systems, all clients see all
writes at the same time. With Transactional Atomicity, clients see
writes at different times depending on which replicas they contact. We
are not aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, Transactional
Atomicity is much weaker than Linearizability.

\noindent There is a substantial \textbf{visibility benefit to sticky
  availability}: writes can safely become visible to readers much more
quickly in a sticky available system. In the model we discussed, it is
possible to achieve several properties like monotonic reads in a
highly available system by waiting until all servers have seen a
write. However, this incurs severe visibility penalties---new writes
will not become visible to clients in the presence of partitions. A
client that does not want to guarantee read-your-writes (due to the
sticky availability requirement) may still wish to read other clients'
writes with timeliness. Accordingly, a system may forgo high
availability in order to provide better visibility. Alternatively, a
system may provide stickiness in the absence of partitions, then
``lose'' session guarantees by having clients reconnect to an
available replica. This latter strategy provides session guarantees in
the absence of partitions but, to provide availability, sacrifices
them in the event of partitions.

>>>>

When possible, we draw on existing properties and definitions from the
database and distributed systems literature, providing a brief,
informal explanation and example for each guarantee. In particular,
for our ACID guarantees, we draw largely on Atul Adya's
dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in our extended Technical Report~\cite{hat-tr}.

@@

To begin, we present achievable semantics, offering proof-of-concept
algorithms to demonstrate feasibility. These guarantees are fairly
straightforward and have been introduced---albeit with greater brevity
and discussion elided---in preliminary prior work~\cite{hat-hotos}. In
our examples, we exclusively consider read and write operations. We
denote a write of value $v$ to data item $d$ as $w_d(v)$ and a read
from data item $d$ returning $v$ as $r_d(v)$ and assume that all data
items have the null value, $\bot$, at database initialization. Unless
otherwise specified, all example transactions commit.

@@

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
al.). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. Clients can buffer their writes until
they commit, or, alternatively, can send them to servers, who will not
serve new writes until clients notify them that the writes have been
committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are promised by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatabable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with
additional properties, like transactional atomicity, it is stronger:
the repeated reads must obey additional ordering constraints. In fact,
distributed systems describe a ``consistent snapshot'' across a set of
related events as a \textit{cut} across a set of participants or, in
our case, data items. Accordingly, to capture the notion that the
transaction should read from a (non-changing) consistent cut over data
items (and to disambiguate from the aforementioned stronger Repeatable
Read properties), we call ANSI SQL Repeatable Read (preventing P2)
\textbf{Cut Isolation}.It is possible to satisfy cut isolation with
high availability by caching appropriate versions for a transaction's
reads or, alternatively, using multi-versioning on the server and
ensuring that each transaction's successive reads return the same
version of each data item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items plus logical ranges of data
items, or predicate-based reads. This prevents the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}. We call predicate-based cut
isolation, which prevents Phantoms, \textit{Phantom Cut
  Isolation}.\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''; this is analogous to Phantom
  Cut Isolation at the level of a single operation within a
  multi-operation transaction. We do not consider this isolation model
  here as it is subsumed by standard Snapshot Isolation and we believe
  our discussion is easily extended to incorporate it.}

@@

Transactional atomicity is core to ACID guarantees. Although, at least
by the acronym, it is not an ``isolation'' property, transactional
atomicity restricts transactions' ability to view the effects of
partially completed transactions. Under transactional atomicity,
within each transaction, either all effects of another transaction are
observed, or none are (equivalently, once some of the effects of a
transaction are observed, all effects are observed). This is a
strictly stronger guarantee than Read Committed isolation: in Read
Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. For example, $T_2$
must observe $b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$.

Perhaps perplexingly, discussions of transactional atomicity are
absent from existing discussions of weak isolation levels. This is
perhaps again due to the single-node context in which prior work was
developed: on a single server (or a fully replicated database),
atomicity is near-trivial to achieve via lightweight locking and/or
local concurrency control over data items. (In this spirit, multi-item
atomicity is common in so-called ``entity groups,'' which logically
represent groups of co-located items~\cite{entitygroup}.) In contrast,
in a distributed environment, atomicity over arbitrary groups of items
is considerably more difficult to achieve with high availability. The
key challenge is that different servers responsible for subsets of a
transaction's updates must coordinate regarding when to reveal their
writes to a transaction---without violating availability due to, say,
locking, or a missing transactional write. Together with cut
isolation, transactional atomicity prevents Read Skew anomalies
(Berenson et al.'s A5A~\cite{ansicritique}).

Transactional atomicity is achievable in a HAT environment. To ensure
transactional atomicity, servers need to know when all of a
transaction's updates are present on their respective
replicas. However, servers do \textit{not} need to agree exactly when
all replicas are aware that all of a transaction's updates are present
on their respective replicas. This subtle distinction corresponds to
the difference between the distributed systems concepts of reliable
broadcast and consensus. The former requires that, if one server
reveals a write, then all servers reveal a write, while the latter
requires explicit agreement across servers (and is unavailable). We
omit a full discussion of the algorithm and instead refer the reader
to the extended version of this paper~\cite{hat-tr}.

@@

So far, our models have not considered what happens across
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). However, in the distributed systems context, many useful
\textit{safety} guarantees span multiple application actions.
\textit{Session guarantees} are used to describe guarantees across
groups of operations (here, transactions) issued by the same logical
user~\cite{sessionguarantees}. Informally, a \textit{session} is an
application context that consists of logical state that should persist
between operations. For example, users accessing a web site may want
to experience a notion of ``forward progress'' for the duration that
they are logged in; all transactions submitted between ``log in'' and
``log out'' operations might form a session. We can define sessions at
several granularities (e.g., per-transaction, per-client, per-time unit)
but, for simplicity, will refer to client-level guarantees in the
discussion.

@@

\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} stipulates
that a client's subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.5em}\noindent\textit{\textbf{Monotonic writes}} requires
that transactions from each individual client be serialized. Any order
on transactions should also respect the order in which each session
submitted transactions.

\vspace{.5em}\noindent\textit{\textbf{Writes Follow Reads}} requires
that, if a session reads a write originating from transaction $T_1$
then performs transaction $T_2$, then a session can only read $T_2$'s
writes if it can also read $T_1$'s prior write (or overwritten values
for $T_1$'s write). This requirement forms the basis for potential
causality: writes follow reads obeys the ``reads-from'' relation that
captures all events that influenced each
transaction~\cite{causalmemory}. Any transaction order should respect
the reads-from order (note that the total order on transactions
precludes reads-from cycles).\vspace{.5em}

@@

\vspace{.5em}\noindent\textit{\textbf{Read your writes}} requires
that whenever a client reads a given data item after updating it, the
read returns the updated value (or a value with a higher ID).

\vspace{.5em}\noindent\textit{\textbf{PRAM}} (Pipelined Random Access
Memory) provides the illusion of serializing each session's operations
and is the combination of monotonic reads, monotonic writes, and read
your writes~\cite{herlihy-art}.

\vspace{.5em}\noindent\textit{\textbf{Causal consistency}} is the combination of
all of the session guarantees~\cite{sessiontocausal} (alternatively,
PRAM and writes-follow-reads) and is also referred to as PL-2L
isolation~\cite{adya}).\vspace{.5em}

@@

\noindent{\textbf{Consistency}} A HAT system can make limited
consistency guarantees. It can execute commutative and logically
monotonic~\cite{calm} operations without the risk of invalidating
(also monotonic) application-level integrity constraints. Our goal in
this paper is not to sketch the entire space of consistency models
that are achievable (see Section~\ref{sec:futurework}, however we
specifically evaluate TPC-C transaction semantics under HAT
consistency guarantees in Section~\ref{sec:evaluation}.

\vspace{.5em}\noindent{\textbf{Durability}} As we briefly discussed in
Section~\ref{sec:availability}, a client requiring durability
of surviving $F$ server faults requires at least ($F+1$)-availability.

\vspace{.5em}\noindent{\textbf{Convergence}} Finally, to
require that the system propagates writes between replicas, we can
require convergence, or eventual consistency for each data item: in
the absence of new mutations to a data item, in the absence of
partitions, all servers eventually agree on the value for each
item. This is typically accomplished by any number of anti-entropy
protocols, which periodically update neighboring servers with the
latest value for each data item. This convergence property should be
considered mandatory given the spirit of existing literature on
isolation protocols: in all models we have encountered, the database
should behave as if there is some total order over
transactions. Indeed, non-convergent systems are remarkably difficult
to reason about.

@@

While there are infinitely many HAT models
(Section~\ref{sec:futurework}), at this point, we have largely
exhausted the range of achievable, previously defined (useful) HAT
semantics. Before summarizing our possibility results, we will present
impossibility results for HATs, also defined in terms of previously
identified isolation and consistency anomalies. Perhaps most notably,
it is impossible to prevent Lost Update or Write Skew in a HAT system.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and serializability---requires unavailability.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)
\end{align*}
The final state of the database will converge to an execution that
could not have resulted from a serial excecution. To prevent this from
happening, either $T_1$ or $T_2$ should not have committed. Each
client's respective server might try to detect that another write
occurred, but this requires knowing the version of the latest write to
$x$. In this example, this reduces to a requirement for
linearizability, which is, via Gilbert and Lynch's proof of the CAP
Theorem, provably unachievable with high
availabilty~\cite{gilbert-cap}.

@@

Data storage engines make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability} or \textit{externally
  consistent transactions} (e.g., Google's
Spanner~\cite{spanner})~\cite{adya}. It is also common, particularly
in systems that allow reading from masters and slaves, to provide a
guarantee such as ``read a version that is no more than five seconds
out of date'' or similar. Unfortunately, an indefinitely long
partition can cause violation of any recency bound, so recency bounds
are unavailable to HAT systems.

@@

We illustrate the hierarchy of available, sticky available, and
unavailble consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable: for example, a HAT system can provide monotonic reads and
transactional atomicity or transactional atomicity and read your
writes and, for maximum semantic richness, can provide all of
transactional atomicity, predicate cut isolation, monotonic reads,
monotonic writes, and writes follow reads. An unavailable system might
provide PRAM and Serializability, as advocated by Salem and
Daudjee~\cite{daudjee-session}.

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Perhaps satisfactorily, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 144 possible highly available
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonimization is \textit{not} exhaustive, but we
believe it lends substantial clarity into the relationships between a
large subset of the prominent ACID and distributed consistency
moels. Additional models of read/write transactions that we have
omitted should be easily classifiable based on the available
primitives and unavailable anomalies we have already discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, Conflict Serializability
(CSR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
\\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

\begin{figure}[t!]
\centering
\begin{tikzpicture}[scale=0.8]
  \tikzstyle{sticky}=[rectangle,draw=blue!50,fill=blue!20,thick]
  \tikzstyle{noha}=[ellipse,draw=red!50,fill=red!20,thick, inner sep=0pt,minimum size=12pt]

@@

\vspace{.5em}\noindent\textbf{Model Composition} Choosing between
combinations of compatible guarantees requires care. Consider the
following transactions:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)
\end{align*}
If we want to guarantee both cut isolation and transactional atomicity
and the system only executes $T_1$, $T_2$, and $T_3$, then $T_3$ needs
to read $a=b=\bot$, $a=b=1$, or $a=b=2$. This means that either the
implementation should frequently return $\bot$ (definitely undesirable
and possibly non-convergent), keep multiple versions of each data item
(necessitating potentially complicated distributed garbage
collection), or use pre-declared read sets to fetch a consistent cut
of keys before each transaction begins to execute. Using client-side
caching can alleviate some of these challenges, but then the system
becomes sticky high available.

Composition cost also varies by property. For instance, Charron-Bost
has proven that, to capture causality between $N$ communicating
processes, standard vector-based approaches face an upper bound of
$O(N)$ storage per write~\cite{charron-bost}. This means that, with
$100K$ clients, each write might be accompanied by $100K$
timestamps per vector. This is difficult to scale. By compromising on
availability (e.g., treating a datacenter as a linearizable cluster),
this overhead can be reduced~\cite{cops, eiger}, but it is much
cheaper to provide, say, read your writes, compared to causality.

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. One might attempt to find a
mapping between the two: transactional atomicity dictates that writes
to multiple keys across multiple servers are made visible to readers
all at once, while linearizability dictates that writes to a single
key on multiple servers are made visible to all readers at once. The
difference is two-old. First, in linearizable (and safe and regular)
systems, writes are made visible to all clients \textit{immediately}
after they finish. With transactional atomicity, there is no recency
guarantee. Second, in linearizabile systems, all clients see all
writes at the same time. With transactional atomicity, clients see
writes at different times depending on which replicas they contact. We
are not aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, transactional
atomicity is much weaker than linearizability.

\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
availability can result in much better write \textit{visibility}:
clients will be able to safely read writes more quickly in a sticky
available system. In the model we discussed, it is possible to achieve
several properties like monotonic reads in a highly available system
by waiting until all servers have seen a write. However, this incurs
severe visibility penalties---new writes will not become visible to
clients in the presence of partitions. A client that does not want to
guarantee read-your-writes (due to the sticky availability
requirement) may still wish to read other clients' writes with
timeliness. Accordingly, a system may forgo high availability in order
to provide better visibility. We also note that a system may wish to
provide stickiness in the absence of partitions, then, in the presence
of partitions, ``lose'' session guarantees by having clients reconnect
to an available replica. This latter strategy provides session
guarantees in the absence of partitions but, to provide availability,
sacrifices them in the event of partitions.
